---
title: "Bioconductor Interactive stats"
abstract: >
  Interactive plots about the packages in Bioconductor project.
date: "`r base::date()`"
output:
  BiocStyle::html_document:
    fig_caption: true
    toc_float:
      collapsed: True
      toc_depth: 3
author:
- name: Lluís Revilla
---
```{r knitrsetup, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_knit$set(root.dir = ".")
knitr::opts_chunk$set(collapse = TRUE, warning = TRUE, fig.wide = TRUE)
BiocStyle::markdown()
options(datatable.print.topn = 10)
library("BiocStyle")
library("ggplot2")
library("data.table")
library("plotly")
library("crosstalk")
```

# Introduction

This page hold interactive plots from the analyis on packages in Bioconductor [^Recommended Chrome but not Firefox]. For static plot one can look at the pages for each category [here](index.html). Many thanks to Carson Sievert for his [guide](https://talks.cpsievert.me/20170125/demo.html)

At the moment for a better interactive plots I am using [ggplot2](https://github.com/tidyverse/ggplot2) and [plotly](https://github.com/ropensci/plotly) from github.

# Select category

Please select category of packages

```{r, echo=FALSE}
load("stats.RData")
stats <- stats[Category == "Experimental", ]
category <- SharedData$new(stats, ~Category, "Selected categories")
stats2 <- SharedData$new(stats, ~Package, "Selected packages")
theme_bw <- theme_bw(base_size = 16)
theme <- theme(axis.text.x = element_text(angle = 60, hjust = 1))
scal <- scale_x_datetime(date_breaks = "3 months")


p <- ggplot(stats2, aes(Date, Nb_of_distinct_IPs, col = Package)) + 
  geom_line() + 
  theme_bw +
  ggtitle("IPs") +
  ylab("Distinct IP downloads") +
  scal +
  theme + 
  guides(col = FALSE)
p %>%
  ggplotly(tooltip = c("Date", "Nb_of_distinct_IPs", "Package"), width = 1200, height = 600) %>%
  highlight(
    on = "plotly_click", 
    off = "plotly_doubleclick", 
    selectize = TRUE,
  )
# 
# ggplot(stats, aes(Date, Nb_of_downloads, col = Package)) + 
#   geom_line() + 
#   theme_bw +
#   ggtitle("Package Downloads") +
#   ylab("Downloads") +
#   scal +
#   theme + 
#   guides(col = FALSE)
# 
# ggplot(stats, aes(Date, Nb_of_downloads/Nb_of_distinct_IPs, col = Package)) + 
#   geom_line() + 
#   theme_bw +
#   ggtitle("Spread of downloads") +
#   ylab("Downloads per IPs") +
#   scal +
#   theme + 
#   guides(col = FALSE)
# PercDate <- stats[, .(Package, Downloads = Nb_of_downloads/sum(Nb_of_downloads)), by = Date]
# PercDate <- PercDate[order(Date, order(Downloads)), ]
# OrdDate <- PercDate[, .(Package, Ord = Downloads/max(Downloads)), by = Date]
# rankDate <- OrdDate[, .(Package, rank = rank(Ord)/.N), by = Date]
# ggplot(rankDate, aes(Date, rank, col = Package)) + 
#   geom_line() + 
#   theme_bw + 
#   ggtitle("Rank of packages by downloads") +
#   xlab("") + 
#   ylab("Position by downloads") +
#   guides(col = FALSE) +
#   scal + 
#   theme
# PercPack <- stats[, .(Date, Downloads = Nb_of_downloads/sum(Nb_of_downloads)), by = Package]
# OrdPack <- PercPack[, .(Date, rank = Downloads/max(Downloads)), by = Package]
# ggplot(OrdPack, aes(Date, rank, col = Package)) + 
#   geom_line() + 
#   theme_bw + 
#   ggtitle("Growth of the packages") +
#   xlab("") + 
#   ylab("Downloads/max(Downloads)") +
#   guides(col = FALSE) +
#   scal + 
#   theme
# 
# model <- rankDate[OrdPack, on = c("Package", "Date")]
# ggplot(model[Package %in% c(packages,"RTools4TB", "SemSim")], aes(i.rank, rank, col = Package)) +
#   geom_point() + 
#   theme_bw + 
#   geom_line() +
#   ylab("Position in Bioconductor") + 
#   xlab("Position in package")
# 
# p <- ggplot(tx) +
#     geom_line(aes(x = date, y = median, group = city))
# 
# p %>%
#     ggplotly(tooltip = "city") %>%
#     highlight(
#         on = "plotly_hover", 
#         off = "plotly_doubleclick", 
#         dynamic = TRUE,
#         persistent = TRUE,
#         selectize = TRUE
#     )
```

