---
title: "Bioconductor stats"
abstract: >
  Analysis of the stats of the Software packages in Bioconductor project,
date: "`r base::date()`"
output:
  BiocStyle::html_document2:
    fig_caption: true
    toc_float:
      collapsed: True
      toc_depth: 3
author:
- name: Lluís Revilla
---
```{r knitrsetup, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_knit$set(root.dir = ".")
knitr::opts_chunk$set(collapse = TRUE, warning = TRUE, fig.wide = TRUE)
BiocStyle::markdown()
options(datatable.print.topn = 10)
library("BiocStyle")
library("ggplot2")
library("data.table")
# library("dtplyr")
# library("scales")
# library("dtplyr")
```

# Load data

Here we are going to analyse the software packages of Bioconductor. See the home of the analysis [here](index.html) were we already transformed the data. From that stats we are going to analyse the software category:

```{r stats}
load("stats.RData", verbose = TRUE)
stats <- stats[Category == "Software", ]
yearly <- yearly[Category == "Software", ]
stats
```

There have been `r length(unique(stats$Package))` Software packages in Bioconductor.

# Packages

## Number

First we explore the number of packages being downloaded by month:
```{r number, fig.cap="Packages in Bioconductor with downloads"}
stats <- stats[Nb_of_downloads != 0, ] # We remove rows of packages with a download in that month.
theme_bw <- theme_bw(base_size = 16)
theme <- theme(axis.text.x=element_text(angle = 60, hjust = 1))
scal <- scale_x_datetime(date_breaks = "3 months")
ggplot(stats[, .(Downloads = .N), by = Date], aes(Date, Downloads)) +
  geom_bar(stat = "identity") + 
  theme_bw +
  ggtitle("Packages downloaded") +
  theme + 
  scal + 
  xlab("")
```
The number of packages being downloaded is increasing with time almost exponentially. Partially explained with the incorporation of new packages
```{r number2, fig.cap="Downloads of packages"}
ggplot(stats[, .(Number = sum(Nb_of_downloads)), by = Date], aes(Date, Number)) +
  geom_bar(stat = "identity") + 
  theme_bw +
  ggtitle("Downloads") +
  scal +
  theme + 
  xlab("")
```
Even if the number of packages increase exponentially, the number of the downloads from 2011 grows linearly with time. Which indicates that each time a software package must compete with more packages to be downloaded.

```{r number3, fig.cap="Downloads of packages per package. The error bar indicates the standard error of the mean."}
pd <- position_dodge(0.1)
ggplot(stats[, .(Number = mean(Nb_of_downloads), 
                  sem = sd(Nb_of_downloads)/sqrt(.N)),
              by = Date], aes(Date, Number)) +
  geom_errorbar(aes(ymin = Number - sem, ymax = Number + sem), 
                width = .1, position = pd) +
  geom_point() + 
  geom_line() +
  theme_bw +
  ggtitle("Downloads") +
  ylab("Mean download for a package") +
  scal +
  theme + 
  xlab("")
```
Here we can appreciate that the number of downloads per package hasn't changed much with time. If something, now there is more dispersion between packages downloads. 

## Incorporations

This might be due to an increase in the usage of packages or that new packages bring more users. We start knowing how many packages has been introduced in Bioconductor each month.

```{r new, fig.cap="New packages"}
today <- base::date()
year <- substr(today, 21, 25)
month <- monthsConvert(substr(today, 5, 7))
incorporation <- stats[ , .SD[which.min(Date)], by = Package, .SDcols = "Date"]
histincorporation <- incorporation[, .(Number = .N), by = Date, ]
ggplot(histincorporation, aes(Date, Number)) + 
  geom_bar(stat="identity") + 
  theme_bw + 
  ggtitle("Packages with first download") +
  scal +
  theme +
  xlab("")
```
We can see that there were more than 350 packages before 2009 in Bioconductor, and since them occasionally there is a raise to 50 new downloads (Which would be new packages being added).
```{r new2, fig.cap="New packages. Zoom on the new downloads of packages after 2009."}
ggplot(histincorporation, aes(Date, Number)) + 
  geom_bar(stat="identity") + 
  theme_bw + 
  ggtitle("Packages with first download") +
  scal +
  theme +
  ylim(c(0, 60)) +
  ylab("New packages") +
  xlab("")
```
We can now observe that for each year there are two spikes of new downloads of packages, usually they are the packages being added for the new release of Bioconductor. 

## Removed

Using a similar procedure we can approximate the packages deprecated and removed each month, although a package could not be downloaded and still included in Bioconductor. In this case we look for the last date a package was downloaded, excluding the current month:

```{r deprecated, fig.cap="Date where a package was last downloaded. Aproximates to the date when packages were removed from Bioconductor."}
deprecation <- stats[, .SD[which.max(Date)], by = Package, .SDcols = c("Date",  "Year", "Month")]
deprecation <- deprecation[Month != month & Year == Year, , .SDcols = "Date"] # Before this month
histDeprecation <- deprecation[, .(Number = .N), by = Date, ]
ggplot(histDeprecation, aes(Date, Number)) + 
  geom_bar(stat = "identity") + 
  theme_bw + 
  ggtitle("Packages without downloads") +
  scal +
  theme + 
  ylab("Last seen packages") + 
  xlab("")
```
Here we can see the packages whose last download was in certain month, assuming that this means they are deprecated. It can happen that a package is no longer downloaded but is still in Bioconductor repository, this would be the reason of the spike to 80 packages as per last month.

We further explore how many time between the incorporation of the package and the last download.
```{r timePackage, fig.cap="Time of packages between first and last download. The red line indicates the mean time in Bioconductor"}
df <- merge(incorporation, deprecation, by = "Package")
 # Transform to years
timeBioconductor <- unclass(df$Date.y-df$Date.x)/(365*60*60*24)
hist(timeBioconductor, main = "Time in Bioconductor", xlab = "Years")
abline(v = mean(timeBioconductor), col = "red")
```
We can see that most deprecated packages are less than a year (I would say around two releases) and some stay on Bioconductor up to 6 years  before being removed. Not surprisingly the number of packages incorporated before 2009 and removed from the repository are `r nrow(df) - nrow(deprecation)` packages. But those packages not removed how do they do in Bioconductor?

# Packages downloads

## Ratio downloads per IP
We can start comparing the number of downloads to how many IPs download each package.
```{r downloadsMonth, fig.cap = "Downloads per IP. The error bars indicate the standard error of the mean.", warning=FALSE}
pd <- position_dodge(0.1)
ggplot(stats[, .(Number = mean(Nb_of_downloads/Nb_of_distinct_IPs),
                 sem = sd(Nb_of_downloads/Nb_of_distinct_IPs)/sqrt(.N)), 
              by = c("Date")], aes(Date, Number)) +
  geom_point() +
  geom_errorbar(aes(ymin = Number - sem, ymax = Number + sem),
                width = .1, position = pd) +
  geom_line() +
  theme_bw +
  ggtitle("Downloads per IP") +
  ylab("Mean downloads per IP") +
  xlab("") + 
  theme +
  scal
```
We can see that usually the number of downloads per IP is around 2, but that there is much variation between the packages. In the points marked in red, the variation is bigger than the mean, this might be due to specific packages being downloaded mostly from the same IP:
```{r ratio}
ratio <- stats[, .(Mean = mean(Nb_of_downloads/Nb_of_distinct_IPs),
                   sem = sd(Nb_of_downloads/Nb_of_distinct_IPs)/sqrt(.N),
                   sd = sd(Nb_of_downloads/Nb_of_distinct_IPs),
                   max = max(Nb_of_downloads/Nb_of_distinct_IPs),
                   min = min(Nb_of_downloads/Nb_of_distinct_IPs)), 
              by = c("Package")]
ratio <- ratio[order(Mean, decreasing = TRUE), ]
ratio$Package <- as.character(ratio$Package)
ratio
```
We can see that the package with more downloads from the same IP is `r Biocpkg(ratio[1, 1])`, followed by, `r Biocpkg(ratio[2, 1])`, `r Biocpkg(ratio[3, 1])` and the forth one is `r Biocpkg(ratio[4, 1])`. We can see that some (`r sum(ratio$max == ratio$min & ratio$min == 1, na.rm = TRUE)`) packages have been downloaded each time from different IP. There are `r sum(ratio$Mean < ratio$sd, na.rm = TRUE)` package with more dispersion than mean download per IP, which suggest that are packages highly downloaded in some specific places. 

I am curious how are the default packages of Bioconductor downloaded, let's see where they are:
```{r bioconductor}
ratio[Package %in% bioc_packages, ]
```
`r Biocpkg("BiocInstaller")` is base package more downloaded per IP, maybe because the is necessary to install the other packages in Bioconductor.

Now we explore if there is some seasons cycles in the downloads, as in figure \@ref(fig:number2) seems to be some cycles. 

## By date

First we can explore the number of IPs per month downloading each package:
```{r date1, fig.cap="Distinct IP per package"}
ggplot(stats, aes(Date, Nb_of_distinct_IPs, col = Package)) + 
  geom_line() + 
  theme_bw +
  ggtitle("IPs") +
  ylab("Distinct IP downloads") +
  scal +
  theme + 
  guides(col = FALSE)
```
As we can see there are two groups of packages at the 2009 years, some with low number of IPs and some with bigger number of IPs. As time progress the number of distinct IPs increases for some packages. 
But is the spread in IPs associated with an increase in downloads? 
```{r date2, fig.cap="Downloads per year"}
ggplot(stats, aes(Date, Nb_of_downloads, col = Package)) + 
  geom_line() + 
  theme_bw +
  ggtitle("Package Downloads") +
  ylab("Downloads") +
  scal +
  theme + 
  guides(col = FALSE)
```
Surprisingly some package have a big outburst of downloads to 400k downloads, others to just 100k downloads. But lets focus on the lower end:
```{r date2b, fig.cap="Downloads per year"}
ggplot(stats, aes(Date, Nb_of_downloads, col = Package)) + 
  geom_line() + 
  theme_bw +
  ggtitle("Downloads per package") +
  ylab("Downloads") +
  scal +
  ylim(0, 50000)+
  theme + 
  guides(col = FALSE)
```
There are many packages close to 0 downloads each month, but most packages has less than 10000 downloads per month:
```{r date2c, fig.cap="Downloads per year"}
ggplot(stats, aes(Date, Nb_of_downloads, col = Package)) + 
  geom_line() + 
  theme_bw+
  ggtitle("Downloads per package") +
  ylab("Downloads") +
  scal +
  ylim(0, 10000)+
  theme + 
  guides(col = FALSE)
```
As we can see, in general the month of the year also influences the number of downloads. So we have that from 2010 the factors influencing the downloads are the year, and the month. 

Maybe there is a relationship between the downloads and the number of IPs per date
```{r date3, fig.cap="Ratio downloads per IP per package"}
ggplot(stats, aes(Date, Nb_of_downloads/Nb_of_distinct_IPs, col = Package)) + 
  geom_line() + 
  theme_bw +
  ggtitle("Spread of downloads") +
  ylab("Downloads per IPs") +
  scal +
  theme + 
  guides(col = FALSE)
```
We can see some packages have occasional raises of downloads per IP. But for small ranges we miss a lot of packages:

```{r date4, fig.cap="Ratio downloads per IP per package", warning=FALSE}
ggplot(stats, aes(Date, Nb_of_downloads/Nb_of_distinct_IPs, col = Package)) + 
  geom_line() + 
  theme_bw +
  ggtitle("IPs") +
  ylab("Ratio") +
  scal +
  theme + 
  guides(col = FALSE) +
  ylim(1, 5)
```
But most of the packages seem to be more or less constant and around 2.

## By year
To measure if the same IP has downloaded the same package in several months we can compare if the IPs per year are closer to the IP per month.
```{r yearly}
staty <- stats[, .(Nb_of_distinct_IPs = sum(Nb_of_distinct_IPs)), by = c("Year", "Package")]
year <- staty[yearly, , on = c("Package", "Year")]
year <- year[, Repeated_IP := Nb_of_distinct_IPs-i.Nb_of_distinct_IPs, by = c("Package", "Year")]
year <- year[, Repeated_IP_per := Repeated_IP/Nb_of_distinct_IPs*100, by = c("Package", "Year")]
ggplot(year, aes(Year, Repeated_IP_per)) + 
  geom_violin() +
  xlab("") +
  ylab("Percentatge of repeated IPs") +
  theme_bw +
  guides(col = FALSE)
```
This gives an idea that most of the IPs don't update all their packages to the newest version of each package, or it could be that as most of the IPs are dynamic they change the measure is not completely reliable 

```{r yearly2}
year2 <- year[, .(m = mean(Repeated_IP_per), sem = sd(Repeated_IP_per)/sqrt(.N)), by = "Year"]
ggplot(year2, aes(Year, m)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = m - sem, ymax = m + sem), 
                width = .1, position = pd) + 
  ylab("Percentatge") +
  xlab("") +
  ggtitle("Mean download per ") + 
  theme_bw
```
We can see however that the percentatge of IPs which download the package quite similar for the packages in Bioconductor. It remains also around 17% of the total IPs. 

# Models
## Position in Bioconductor
We can observe if the packages has been consistently the most downloaded package of Bioconductor:
```{r normalize1, fig.cap = "Downloads of packages in Bioconductor. "}
PercDate <- stats[, .(Package, Downloads = Nb_of_downloads/sum(Nb_of_downloads)), by = Date]
PercDate <- PercDate[order(Date, order(Downloads)), ]
ggplot(PercDate, aes(Date, Downloads, col = Package)) + 
  geom_line() + 
  theme_bw + 
  ggtitle("Downloads of packages by dates") +
  xlab("") + 
  ylab("% of downloads") +
  guides(col = FALSE) +
  scal + 
  theme
```
We can see that usually the percentage of downloads of a packages doesn't reach 5% of downloads of the category. And that there is a huge differences between packages in the number of downloads. Let's compare the most downloaded package to the other package:
```{r normalize1b, fig.cap = "Position of packages in Bioconductor"}
OrdDate <- PercDate[, .(Package, Ord = Downloads/max(Downloads)), by = Date]
ggplot(OrdDate, aes(Date, Ord, col = Package)) + 
  geom_line() + 
  theme_bw + 
  ggtitle("Downloads of packages by dates") +
  xlab("") + 
  ylab("% of the Package more downloaded") +
  guides(col = FALSE) +
  scal + 
  theme
```
We can observe that usually there are few package that are closer to the most downloaded packages and lots of them are far from it. We can see the rank between them to be able to see the evolution of downloads along time of a package:

```{r normalize2, fig.cap = "% of packages in Bioconductor. Closer to 1 indicates the top downloaded packages."}
rankDate <- OrdDate[, .(Package, rank = rank(Ord)/.N), by = Date]
ggplot(rankDate, aes(Date, rank, col = Package)) + 
  geom_line() + 
  theme_bw + 
  ggtitle("Rank of packages by downloads") +
  xlab("") + 
  ylab("Position by downloads") +
  guides(col = FALSE) +
  scal + 
  theme
```
Only if we select to follow some package we can track them:
```{r example1, fig.cap="Evolution of downloads of packages in Bioconductor"}
packages <- c("limma", "GOSemSim", "BioCor", "Clonality", "Prostar", 
             "rintact", "bioassayR", "DESeq", "DESeq2", "edgeR")
ggplot(rankDate[Package %in% packages, ], aes(Date, rank, col = Package))+ 
  geom_line() + 
  theme_bw + 
  ggtitle("Relative rank of packages by downloads") +
  xlab("") + 
  ylab("Position by downloads") +
  scal + 
  theme
```
Here we can see that limma is one of the top downloaded packages since 2009, `r Biocpkg("edgeR")` and `r Biocpkg("GOSemSim")` has grown to reach the top downloaded packages, while other packages haven't reached the top 25% packages by downloads on a month. 

## Package cycle

We can also observe when did a package reach the maximum number of downloads:
```{r cycle1, fig.cap = "Cycle of packages. Percentatge of downloads of a package along time."}
PercPack <- stats[, .(Date, Downloads = Nb_of_downloads/sum(Nb_of_downloads)), by = Package]
ggplot(PercPack, aes(Date, Downloads, col = Package)) + 
  geom_line() + 
  theme_bw + 
  ggtitle("Growth of the packages") +
  xlab("") + 
  ylab("Downloads/max(Downloads)") +
  guides(col = FALSE) +
  scal + 
  theme
```
Here we can see the date when a package reached the highest number of downloads. For most packages the higher downloads are on the recent months.
```{r cycle2a, fig.cap = ""}
OrdPack <- PercPack[, .(Date, rank = Downloads/max(Downloads)), by = Package]
ggplot(OrdPack, aes(Date, rank, col = Package)) + 
  geom_line() + 
  theme_bw + 
  ggtitle("Growth of the packages") +
  xlab("") + 
  ylab("Downloads/max(Downloads)") +
  guides(col = FALSE) +
  scal + 
  theme
```
Here we can see when the package hast the most downloads along time. As usually we need to focus on fewer packages to be able to distinguish between package to see their evolution:
```{r cycl2, fig.cap = "Cycle of few packages. Position of package downloads respect the maximum downloads of the packages along time."}
ggplot(OrdPack[Package %in% c(packages, "RTools4TB", "SemSim"), ], 
       aes(Date, rank, col = Package)) + 
  geom_line() + 
  theme_bw + 
  ggtitle("Growth of the packages") +
  xlab("") + 
  ylab("Downloads/max(Downloads)") +
  scal + 
  theme
```
As expected the packages that keep up with Bioconductor growth have a peak near the end of the series. For this reason I added the package `r Biocpkg("SemSim")` and the `r Biocpkg("RTools4TB")` to see that package that has been less and less downloaded.

We can combine both models into a single one to have a timeless comparison of the packages. So we will know if the max position reached in Bioconductor is done when there are more  downloads. 
```{r model, fig.cap="Relative position of the packages. Position in Bioconductor downloads and position of package downloads."}
model <- rankDate[OrdPack, on = c("Package", "Date")]
setnames(model, "rank", "rank.B")
setnames(model, "i.rank", "rank.P")
ggplot(model[Package %in% c(packages,"RTools4TB", "SemSim")], aes(rank.P, rank.B, col = Package)) +
  geom_point() + 
  theme_bw + 
  geom_line() +
  ylab("Position in Bioconductor") + 
  xlab("Position in package")
```
This is a timeless comparison of the growth of packages in Bioconductor and against themselves. The higher on the y-axis the higher they have been on Bioconductor, and the higher on the x-axis the most downloads they had. 
Next we add the time as a factor so we standarize by the time a package has been in Bioconductor. We have omitted months with 0 downloads of a package so this might bias a bit the ranking.
```{r model2, fig.cap="Relative position of the packages. Position relative to Bioconductor and itself compared to date."}
model[, rank.D := rank(as.numeric(Date))/.N, by = Package]
ggplot(model[Package %in% c(packages,"RTools4TB", "SemSim")], aes(rank.D, rank.B, col = Package)) +
  geom_point() + 
  theme_bw + 
  geom_line() +
  xlab("By date") + 
  ylab("Position in Bioconductor")
  
ggplot(model[Package %in%  c(packages,"RTools4TB", "SemSim")], aes(rank.D, rank.P, col = Package)) +
  geom_point() + 
  theme_bw + 
  geom_line() +
  xlab("By date") + 
  ylab("Position in Package")
```
We can observe that some packages climb to the top of Bioconductor sooner than others and some climb and then decrease the number of downloads proportional to the the number of downloads in Bioconductor.

# SessionInfo {.unnumbered}

```{r sessioninfo}
sessionInfo()
```

