---
title: "Bioconductor stats"
abstract: >
  Analysis of the stats of the support site of Bioconductor project.
date: "`r base::date()`"
output:
  BiocStyle::html_document2:
    fig_caption: true
    toc_float:
      collapsed: True
      toc_depth: 3
author:
- name: Lluís Revilla
---
```{r knitrsetup, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_knit$set(root.dir = ".")
knitr::opts_chunk$set(collapse = TRUE, warning = TRUE, fig.wide = TRUE)
BiocStyle::markdown()
options(datatable.print.topn = 10)
library("BiocStyle")
library("ggplot2")
library("data.table")
library("jsonlite")
# library("scales")
```

# Introduction

Here we are going to analyse the usage of the support site of Bioconductor and its relationship with the packages. See the home of the analysis [here](index.html).

We will use the api of the [site](https://support.bioconductor.org/info/api/).

# First post


```{r api0}
base <- "https://support.bioconductor.org/api/"
post0 <- fromJSON(paste0(base, "stats/day/0"))
post0$date
post0$type
```
Here we can see that the first post in bioconductor is from 2001, but it must be from the mailing list because it doesn't have a type description.

# First day 

The first day of the support site was 16/09/2014 as you can see [here](https://groups.google.com/forum/#!topic/rbioc-sig-proteomics/3Bmnzx5nsak)
```{r firstDay}
post1 <- fromJSON(paste0(base, "stats/date/2014/09/16"))
(original_posts <- sum(post1$answers, post1$comments, post1$questions))
```


# Today
Total amount of post.
```{r today}
yesterday <- gsub("-", "/", Sys.Date()-1) # Upt to yesterday
posts <- fromJSON(paste0(base, "stats/date/", yesterday))
(total_posts <- sum(posts$answers, posts$comments, posts$questions))
```

# Every year
We can see the participation for each year
```{r}
endYear <- paste0(2014:2016, "/12/31")
yearPosts <- sapply(endYear, function(x){
  posts <- fromJSON(paste0(base, "stats/date/", x))
  sum(posts$answers, posts$comments, posts$questions) - original_posts
})
barplot(yearPosts)
```
We can se an increase of the participation (it doesn't count news or Jobs though)

# Questions
To retrieve the questions posted for each package first we need to know the questions:
```{r questions}
rep <- round((total_posts-original_posts)/1000) + 3 # To count news, jobs...
questions <- sapply(seq_len(rep), function(i){
  ipost <- original_posts + (i-1)*1000
  fpost <- original_posts + i*1000
  message("Quering from ", ipost, " to ", fpost)
  ids <- seq(from = ipost, to = fpost)
  posts <- sapply(as.character(ids), function(x) {
    try(p <- fromJSON(paste0(base, "post/", x, "/")), silent = TRUE)
    if (exists("p")) {
      if (p$type == "Question") {
        p
      }
    }
  }, simplify = FALSE)
  
  message("Questions retrieved")
  Sys.sleep(1)
})
questions <- questions[lenghts(questions) == 26]
save(questions, yesterday, file = "questions.RData")
```


